MAESTROeX: A Massively Parallel Low Mach Number Astrophysical Solver

### 2019-06-07 MAESTROeX_I/intro.tex
sec 1 updates

Low Mach number models have been developed for a variety of contexts
including combustion \citep{day2000numerical}, terrestrial atmospheric
modeling \citep{durran:1989,oneill:2014,duarte2015low}, and elastic
solids \citep{abbate2017all}.  For astrophysical applications, a
number of approaches to modeling low Mach number flows have been
developed in recent years.  These include methods atmospheric model similar to
what is presented here~\cite{Lin:2006}.  There are also  semi-implicit all-Mach
number solvers, where the Euler equations are split into an acoustic
part and an advective part
\citep{Kwatra2009,Degond2009,Cordier2012,Haack2012,Happenhofer2013,Chalons2016,Padioleau2019}.
The fast acoustic waves are then solved using implicit time
integration, while the slow material waves are solved explicitly.
Another approach is to use preconditioned all-Mach number solvers
\citep{Miczek2014,Barsukow2016}, where the numerical flux is
multiplied by a preconditioning matrix.  This reduces the stiffness of
the system at low Mach numbers, while retaining the correct scaling
behavior. In the reduced speed of sound technique (RSST) and related
methods, the speed of sound is artificially reduced by including a
suitable scaling factor in the continuity equation, reducing the
restriction on the size of the timestep
\citep{Rempel2005,Hotta2012,Wang2015,Takeyama2017,Iijima2018}.
Finally, there are fully implicit time integration codes for the
compressible Euler equations
\citep{Viallet2011,kifonidis:2012,Viallet2015,Goffrey2016}.  The MUSIC code
uses fully implicit time integration for the compressible Euler
equations, which therefore allows for arbitrarily large timesteps.

Previously, we developed the low Mach number astrophysical solver, MAESTRO.
MAESTRO is a block-structured, Cartesian grid finite-volume, adaptive mesh refinement (AMR) code that has been successfully used for many years for a number of applications, detailed below.
Unlike several of the references above, MAESTRO is not an all-Mach solver, but is suitable for flows where the Mach number is small ($\sim 0.1$ or smaller).  
Furthermore, the low Mach number model in MAESTRO is specifically designed for, but not limited to, astrophysical settings with significant atmospheric stratification.
This includes full spherical stars, as well as planar simulations of dynamics within localized regions of a star.
MAESTRO uses an explicit Godunov approach for advection, a stiff ODE package (VODE) \citep{vode} for reactions, and multigrid-based linear solvers for the pressure-projection steps, and is thus time-step limited by an advective CFL constraint based on the fluid velocity, not the sound speed.
The time-advancement strategy uses Strang splitting to couple the thermodynamic variables, and a second-order projection method to integrate the velocity equations subject to a divergence constraint.
Central to the algorithm is a time-varying, one-dimensional stratified background (or base) state density and pressure held in hydrostatic equilibrium that couple to the full Cartesian grid solution.
The original MAESTRO code was developed in the pure-Fortran 90 FBoxLib software framework (available in the same github organization as AMReX).

@@

In this paper, we present new algorithmic methodology that improves upon Paper V in a number of ways.
First, the overall temporal algorithm has been greatly simplified without compromising accuracy.
The key design decisions were to eliminate the splitting of the velocity into average and perturbational components, 
and also to replace the hydrodynamic evolution of the base state with a predictor-corrector approach.
Not only does this greatly simplify the dynamics of the base
state, but this treatment is more amenable to higher-order multiphysics coupling strategies
based on method-of-lines integration.  
In particular, schemes based on spectral deferred corrections have been used to generate high-order temporal integrators for problems of reactive flow \citep{dutt2000spectral} and low Mach number combustion \citep{pazner2016high,nonaka2018conservative}.
Second, we explore the effects of alternative spatial mapping routines for coupling the base state and the Cartesian grid state in spherical problems.
Finally, the code has been completely ported to a new software framework suitable for modern manycore and GPU-based supercomputers.
We examine the performance of our new MAESTROeX implementation in the new C++/F90 AMReX public software library \citep{AMReX,AMReX_JOSS}.
MAESTROeX uses MPI+OpenMP parallelism and scales well to over 10,000 MPI processes, with each MPI process supporting tens of threads.
The resulting code is publicly available on GitHub ({\tt https://github.com/AMReX-Astro/MAESTROeX}),
uses the Starkiller-Astro microphysics libraries \citep{starkiller},
and the AMReX software library for block-structured adaptive mesh refinement ({\tt https://github.com/AMReX-Codes/amrex}).

>>>>

Low Mach number models have been developed for a variety of contexts
including combustion \citep{day2000numerical}, terrestrial atmospheric
modeling \citep{durran:1989,oneill:2014,duarte2015low}, and elastic
solids \citep{abbate2017all}.  For astrophysical applications, a
number of approaches to modeling low Mach number flows have been
developed in recent years.  One of the more similar approaches to ours is
\cite{Lin:2006}, however this approach is only first-order accurate and does
not account for atmospheric expansion.  There are also semi-implicit all-Mach
number solvers, where the Euler equations are split into an acoustic
part and an advective part
\citep{Kwatra2009,Degond2009,Cordier2012,Haack2012,Happenhofer2013,Chalons2016,Padioleau2019}.
The fast acoustic waves are then solved using implicit time
integration, while the slow material waves are solved explicitly.
Another approach is to use preconditioned all-Mach number solvers
\citep{Miczek2014,Barsukow2016}, where the numerical flux is
multiplied by a preconditioning matrix.  This reduces the stiffness of
the system at low Mach numbers, while retaining the correct scaling
behavior. In the reduced speed of sound technique (RSST) and related
methods, the speed of sound is artificially reduced by including a
suitable scaling factor in the continuity equation, reducing the
restriction on the size of the timestep
\citep{Rempel2005,Hotta2012,Wang2015,Takeyama2017,Iijima2018}.
Finally, there are fully implicit time integration codes for the
compressible Euler equations
\citep{Viallet2011,kifonidis:2012,Viallet2015,Goffrey2016}.  The MUSIC code
uses fully implicit time integration for the compressible Euler
equations, which therefore allows for arbitrarily large timesteps.

Previously, we developed the low Mach number astrophysical solver, MAESTRO.
MAESTRO is a block-structured, Cartesian grid finite-volume, adaptive mesh refinement (AMR) 
code that has been successfully used for many years for a number of applications, detailed below.
Unlike several of the references above, MAESTRO is not an all-Mach solver, but is suitable for 
flows where the Mach number is small ($\sim 0.1$ or smaller).  
Furthermore, the low Mach number model in MAESTRO is specifically designed for, but not limited 
to, astrophysical settings with significant atmospheric stratification.
This includes full spherical stars, as well as planar simulations of dynamics within localized 
regions of a star.
The numerical methodology relies on an explicit Godunov approach for advection, a stiff ODE 
solver for reactions (VODE) \citep{vode}, and multigrid-based linear solvers for the 
pressure-projection steps.  This, the time step is limited by an advective CFL constraint based on the 
fluid velocity, not the sound speed.
Central to the algorithm are time-varying, one-dimensional stratified background (or base) state 
density and pressure fields that are held in hydrostatic equilibrium.
The base state density couples to the full state solution through buyoancy terms in the momentum equation,
and the base state pressure couples to the full state solution by constraining the evolution of the
thermodynamic variables to match this pressure.
The time-advancement strategy uses a predictor-corrector strategy for evolving the base state,
Strang splitting to integrate the thermodynamic variables, and a 
second-order projection method to integrate the velocity equations subject to a divergence constraint.
The original MAESTRO code was developed in the pure-Fortran 90 FBoxLib software framework, whereas
MAESTROeX is developed in the C++/F90 AMReX framework \citep{AMReX,AMReX_JOSS}.

@@

In this paper, we present new algorithmic methodology that improves upon Paper V in a number of ways.
First, the overall temporal algorithm has been greatly simplified without compromising second-order accuracy.
The key design decisions were to eliminate the splitting of the velocity into average and perturbational components, 
and also to replace the hydrodynamic evolution of the base state with a predictor-corrector approach.
Not only does this greatly simplify the dynamics of the base
state, but this treatment is more amenable to higher-order multiphysics coupling strategies
based on method-of-lines integration.  
In particular, schemes based on deferred corrections have been used to generate high-order temporal integrators for problems of reactive flow \citep{dutt2000spectral} and low Mach number combustion \citep{pazner2016high,nonaka2018conservative}.
Second, we explore the effects of alternative spatial mapping routines for coupling the base state and the Cartesian grid state for spherical problems.
Finally, we examine the performance of our new MAESTROeX implementation in the new C++/F90 AMReX public software library \citep{AMReX,AMReX_JOSS}.
MAESTROeX uses MPI+OpenMP parallelism and scales well to over 10,000 MPI processes, with each MPI process supporting tens of threads.
The resulting code is publicly available on GitHub ({\tt https://github.com/AMReX-Astro/MAESTROeX}),
uses the Starkiller-Astro microphysics libraries \citep{starkiller} ({\tt https://github.com/starkiller-astro}) ,
as well as AMReX ({\tt https://github.com/AMReX-Codes/amrex}).

### 2019-06-07 MAESTROeX_I/paper.tex
Added figures to appendix A
sec 1 updates

@@

%% Mark off the abstract in the ``abstract'' environment.
\begin{abstract}
We present MAESTROeX, a massively parallel solver for low Mach number astrophysical flows.
The underlying low Mach number equation set allows for efficient, long-time integration for highly subsonic flows compared to compressible approaches.
MAESTROeX is suitable for modeling full spherical stars as well as well as planar simulations of dynamics within localized regions of a star,
and can robustly handle several orders of magnitude of density and pressure stratification.
Previously, we have described the development of the predecessor of MAESTROeX, called MAESTRO, in a series of papers. 
Here, we present a new, greatly simplified temporal integration scheme that retains the same order of accuracy as our previous approaches. 
We also explore the use of alternative spatial mapping of the one-dimensional base state onto the full Cartesian grid.
The code leverages the new AMReX software framework for block-structured adaptive mesh refinement applications, allowing for scalability to large fractions of leadership-class machines.
Using our previous studies on the convective phase of single-degenerate progenitor models of Type Ia supernovae as a guide, we characterize the performance of the code and validate the new algorithmic features.
\end{abstract}

>>>>

\usepackage{multirow}

@@

%% Mark off the abstract in the ``abstract'' environment.
\begin{abstract}
We present MAESTROeX, a massively parallel solver for low Mach number astrophysical flows.
The underlying low Mach number equation set allows for efficient, long-time integration for highly subsonic flows compared to compressible approaches.
MAESTROeX is suitable for modeling full spherical stars as well as well as planar simulations of dynamics within localized regions of a star,
and can robustly handle several orders of magnitude of density and pressure stratification.
Previously, we have described the development of the predecessor of MAESTROeX, called MAESTRO, in a series of papers. 
Here, we present a new, greatly simplified temporal integration scheme that retains the same order of accuracy as our previous approaches. 
We also explore the use of alternative spatial mapping of the one-dimensional base state onto the full Cartesian grid.
The code leverages the new AMReX software framework for block-structured adaptive mesh refinement (AMR) applications, allowing for scalability to large fractions of leadership-class machines.
Using our previous studies on the convective phase of single-degenerate progenitor models of Type Ia supernovae as a guide, we characterize the performance of the code and validate the new algorithmic features.  Like MAESTRO, MAESTROeX is fully open source.
\end{abstract}
